---
title: "Turnaround Time Report"
subtitle: "Weekly TAT Analysis by Examination Category"
author: "Lab Intelligence"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(scales)
library(gt)

theme_set(theme_minimal(base_size = 12))

# Category labels
category_labels <- c(
 "101" = "101 - Culture",
 "221" = "221 - PCR",
 "218" = "218 - POCT PCR"
)
```

```{r}
#| label: load-data
# Load data (generate first if needed)
if (!file.exists("tat_data.rds")) {
 source("generate_data.R")
}
tat_data <- readRDS("tat_data.rds")

# Calculate TAT segments (in minutes)
tat_analysis <- tat_data |>
 mutate(
   # Segment 1: Pre-lab (afsendt -> modtaget)
   seg1_prelab = as.numeric(difftime(modtaget, afsendt, units = "mins")),
   # Segment 2: In-lab (modtaget -> besvaret)
   seg2_inlab = as.numeric(difftime(besvaret, modtaget, units = "mins")),
   # Segment 3: Post-lab (besvaret -> kvitteret)
   seg3_postlab = as.numeric(difftime(kvitteret, besvaret, units = "mins")),
   # Segment 4: End-to-end lab (afsendt -> besvaret)
   seg4_e2e_lab = as.numeric(difftime(besvaret, afsendt, units = "mins")),
   # Segment 5: Total clinical (afsendt -> kvitteret)
   seg5_total = as.numeric(difftime(kvitteret, afsendt, units = "mins")),
   # Week for grouping
   week = floor_date(afsendt, "week"),
   kategori_label = category_labels[kategori]
 )
```

## Overview

This report analyzes turnaround times across `r n_distinct(tat_data$kategori)` examination categories over `r n_distinct(floor_date(tat_data$afsendt, "week"))` weeks.

```{r}
#| label: tbl-overview
#| tbl-cap: "Sample counts by category"
tat_analysis |>
 group_by(Category = kategori_label) |>
 summarise(
   `Total Samples` = n(),
   `Pending Result` = sum(is.na(besvaret)),
   `Pending Acknowledgment` = sum(is.na(kvitteret) & !is.na(besvaret)),
   .groups = "drop"
 ) |>
 gt() |>
 tab_header(title = "Sample Volume Summary")
```

## TAT Segments

The turnaround time is broken into five segments:
| Segment | Definition | Measures |
|---------|------------|----------|
| 1: Pre-lab | Requisition → Received | Collection & transport time |
| 2: In-lab | Received → Released | Laboratory processing time |
| 3: Post-lab | Released → Acknowledged | Clinical uptake time |
| 4: End-to-end | Requisition → Released | Total lab responsibility |
| 5: Total | Requisition → Acknowledged | Complete clinical cycle |


## Process Flow: Median TAT by Category

```{r}
#| label: fig-process-flow
#| fig-cap: "Median TAT by segment and category (completed samples only)"
#| fig-height: 5
# Calculate medians for completed samples
flow_data <- tat_analysis |>
 filter(!is.na(kvitteret)) |>
 group_by(kategori_label) |>
 summarise(
   seg1 = median(seg1_prelab, na.rm = TRUE),
   seg2 = median(seg2_inlab, na.rm = TRUE),
   seg3 = median(seg3_postlab, na.rm = TRUE),
   .groups = "drop"
 ) |>
 pivot_longer(cols = starts_with("seg"), names_to = "segment", values_to = "minutes") |>
 mutate(
   segment_label = case_match(
     segment,
     "seg1" ~ "Pre-lab",
     "seg2" ~ "In-lab",
     "seg3" ~ "Post-lab"
   ),
   segment_label = factor(segment_label, levels = c("Pre-lab", "In-lab", "Post-lab"))
 )

# Format time helper
format_time <- function(mins) {
 if(is.na(mins)) return(NA_character_)
 if (mins < 60) {
   paste0(round(mins), "m")
 } else if (mins < 24 * 60) {
   paste0(round(mins / 60, 1), "h")
 } else {
   paste0(round(mins / (24 * 60), 1), "d")
 }
}

ggplot(flow_data, aes(x = segment_label, y = minutes, fill = kategori_label)) +
 geom_col(position = "dodge", width = 0.7) +
 geom_text(
   aes(label = sapply(minutes, format_time)),
   position = position_dodge(width = 0.7),
   vjust = -0.5,
   size = 3
 ) +
 scale_fill_brewer(palette = "Set2", name = "Category") +
 scale_y_continuous(
   labels = function(x) sapply(x, format_time),
   expand = expansion(mult = c(0, 0.15))
 ) +
 labs(
   x = "TAT Segment",
   y = "Median Time",
   title = "Process Flow: Median TAT by Segment",
   subtitle = "Requisition → Received → Released → Acknowledged"
 ) +
 theme(legend.position = "bottom")
```

## Percentile Distribution: In-Lab TAT (Segment 2)

```{r}
#| label: tbl-percentiles-inlab
#| tbl-cap: "In-lab TAT percentiles by category"
percentiles <- c(0.5, 0.8, 0.9, 0.95, 0.99)

inlab_percentiles <- tat_analysis |>
 filter(!is.na(seg2_inlab)) |>
 group_by(Category = kategori_label) |>
 summarise(
   P50 = quantile(seg2_inlab, 0.50),
   P80 = quantile(seg2_inlab, 0.80),
   P90 = quantile(seg2_inlab, 0.90),
   P95 = quantile(seg2_inlab, 0.95),
   P99 = quantile(seg2_inlab, 0.99),
   .groups = "drop"
 )

inlab_percentiles |>
 gt() |>
 tab_header(
   title = "In-Lab TAT (Segment 2): Percentile Distribution",
   subtitle = "Time from sample received to result released"
 ) |>
 fmt_number(columns = -Category, decimals = 0, pattern = "{x} min") |>
 cols_align(align = "right", columns = -Category)
```

```{r}
#| label: fig-percentiles-inlab
#| fig-cap: "In-lab TAT percentile distribution by category"
#| fig-height: 5
percentile_long <- inlab_percentiles |>
 pivot_longer(-Category, names_to = "Percentile", values_to = "minutes") |>
 mutate(Percentile = factor(Percentile, levels = c("P50", "P80", "P90", "P95", "P99")))

ggplot(percentile_long, aes(x = Percentile, y = minutes, color = Category, group = Category)) +
 geom_line(linewidth = 1) +
 geom_point(size = 3) +
 scale_color_brewer(palette = "Set2") +
 scale_y_continuous(labels = function(x) sapply(x, format_time)) +
 labs(
   x = "Percentile",
   y = "Time",
   title = "In-Lab TAT Percentile Distribution"
 ) +
 theme(legend.position = "bottom")
```

## All Segments: Percentile Tables

::: {.panel-tabsettat-re}

### Pre-lab (Segment 1)
```{r}
#| label: tbl-percentiles-prelab
tat_analysis |>
 filter(!is.na(seg1_prelab)) |>
 group_by(Category = kategori_label) |>
 summarise(
   P50 = quantile(seg1_prelab, 0.50),
   P80 = quantile(seg1_prelab, 0.80),
   P90 = quantile(seg1_prelab, 0.90),
   P95 = quantile(seg1_prelab, 0.95),
   P99 = quantile(seg1_prelab, 0.99),
   .groups = "drop"
 ) |>
 gt() |>
 tab_header(title = "Pre-lab TAT (Requisition → Received)") |>
 fmt_number(columns = -Category, decimals = 0, pattern = "{x} min")
```

### In-lab (Segment 2)
```{r}
#| label: tbl-percentiles-inlab-tab
inlab_percentiles |>
 gt() |>
 tab_header(title = "In-lab TAT (Received → Released)") |>
 fmt_number(columns = -Category, decimals = 0, pattern = "{x} min")
```

### Post-lab (Segment 3)
```{r}
#| label: tbl-percentiles-postlab
tat_analysis |>
 filter(!is.na(seg3_postlab)) |>
 group_by(Category = kategori_label) |>
 summarise(
   P50 = quantile(seg3_postlab, 0.50),
   P80 = quantile(seg3_postlab, 0.80),
   P90 = quantile(seg3_postlab, 0.90),
   P95 = quantile(seg3_postlab, 0.95),
   P99 = quantile(seg3_postlab, 0.99),
   .groups = "drop"
 ) |>
 gt() |>
 tab_header(title = "Post-lab TAT (Released → Acknowledged)") |>
 fmt_number(columns = -Category, decimals = 0, pattern = "{x} min")
```

### End-to-end (Segment 4)
```{r}
#| label: tbl-percentiles-e2e
tat_analysis |>
 filter(!is.na(seg4_e2e_lab)) |>
 group_by(Category = kategori_label) |>
 summarise(
   P50 = quantile(seg4_e2e_lab, 0.50),
   P80 = quantile(seg4_e2e_lab, 0.80),
   P90 = quantile(seg4_e2e_lab, 0.90),
   P95 = quantile(seg4_e2e_lab, 0.95),
   P99 = quantile(seg4_e2e_lab, 0.99),
   .groups = "drop"
 ) |>
 gt() |>
 tab_header(title = "End-to-end Lab TAT (Requisition → Released)") |>
 fmt_number(columns = -Category, decimals = 0, pattern = "{x} min")
```

### Total (Segment 5)
```{r}
#| label: tbl-percentiles-total
tat_analysis |>
 filter(!is.na(seg5_total)) |>
 group_by(Category = kategori_label) |>
 summarise(
   P50 = quantile(seg5_total, 0.50),
   P80 = quantile(seg5_total, 0.80),
   P90 = quantile(seg5_total, 0.90),
   P95 = quantile(seg5_total, 0.95),
   P99 = quantile(seg5_total, 0.99),
   .groups = "drop"
 ) |>
 gt() |>
 tab_header(title = "Total Clinical TAT (Requisition → Acknowledged)") |>
 fmt_number(columns = -Category, decimals = 0, pattern = "{x} min")
```

:::

## Weekly Trend: In-Lab TAT (P90)

```{r}
#| label: fig-trend
#| fig-cap: "Weekly trend of In-lab TAT (90th percentile)"
#| fig-height: 5
weekly_trend <- tat_analysis |>
 filter(!is.na(seg2_inlab)) |>
 group_by(week, kategori_label) |>
 summarise(
   p90 = quantile(seg2_inlab, 0.90),
   n = n(),
   .groups = "drop"
 )

ggplot(weekly_trend, aes(x = week, y = p90, color = kategori_label)) +
 geom_line(linewidth = 1) +
 geom_point(size = 3) +
 scale_color_brewer(palette = "Set2", name = "Category") +
 scale_x_datetime(date_labels = "%b %d", date_breaks = "1 week") +
 scale_y_continuous(labels = function(x) sapply(x, format_time)) +
 labs(
   x = "Week",
   y = "P90 In-lab TAT",
   title = "In-Lab TAT Trend (90th Percentile)",
   subtitle = "Weekly performance by category"
 ) +
 theme(legend.position = "bottom")
```

## Distribution: In-Lab TAT

```{r}
#| label: fig-distribution
#| fig-cap: "Distribution of In-lab TAT by category"
#| fig-height: 5
ggplot(
 tat_analysis |> filter(!is.na(seg2_inlab)),
 aes(x = seg2_inlab, fill = kategori_label)
) +
 geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
 facet_wrap(~kategori_label, scales = "free", ncol = 1) +
 scale_fill_brewer(palette = "Set2", guide = "none") +
 scale_x_continuous(labels = function(x) sapply(x, format_time)) +
 labs(
   x = "In-lab TAT",
   y = "Count",
   title = "In-Lab TAT Distribution by Category"
 )
```

## Pending Samples

```{r}
#| label: tbl-pending
#| tbl-cap: "Samples pending result or acknowledgment"
pending <- tat_analysis |>
 filter(is.na(besvaret) | is.na(kvitteret)) |>
 mutate(
   status = case_when(
     is.na(besvaret) ~ "Pending result",
     is.na(kvitteret) ~ "Pending acknowledgment"
   ),
   age_hours = as.numeric(difftime(Sys.time(), afsendt, units = "hours"))
 ) |>
 select(proevenr, kategori_label, status, afsendt, age_hours)

if (nrow(pending) > 0) {
 pending |>
   arrange(desc(age_hours)) |>
   gt() |>
   tab_header(title = "Pending Samples") |>
   fmt_datetime(columns = afsendt, date_style = "iso", time_style = "iso") |>
   fmt_number(columns = age_hours, decimals = 1, pattern = "{x}h") |>
   cols_label(
     proevenr = "Sample ID",
     kategori_label = "Category",
     status = "Status",
     afsendt = "Requisition Time",
     age_hours = "Age"
   )
} else {
 cat("No pending samples.")
}
```

